---
title: "Descriptive Statistics"
author: "Allison McCarty"
date: "4/11/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)
library(readxl)
library(tidyverse)
library(dplyr)
library(pastecs)
library(data.table)
library(formattable)
library(tidyr)
library(mice)
library(tibble)
library(GGally)
library(kableExtra)
library(ggpubr)
library(zoo)


# Read in Data
otter <- read.csv("~/Downloads/otter2.csv")

#Check for NA's
sum(apply(otter, 1, function(x){any(is.na(x))})) #84/100 observations have at least one missing value
sum(is.na(otter$Otter)) # NONE
sum(is.na(otter$Activity)) #NONE
sum(is.na(otter$Altitude)) #18 MISSING VALUES
sum(is.na(otter$Trout)) #16 MISSING VALUES
sum(is.na(otter$Site)) #NONE
sum(is.na(otter$Month)) #NONE
sum(is.na(otter$Lat.)) #NONE
sum(is.na(otter$Long.)) # NONE
sum(is.na(otter$Caloric)) #NONE
sum(is.na(otter$Water.Excess)) #NONE
sum(is.na(otter$Amoeba)) #40 MISSING VALUES
sum(is.na(otter$Land)) #NONE
sum(is.na(otter$Water)) #40 MISSING VALUES
sum(is.na(otter$Month.1)) #NONE
sum(is.na(otter$Year)) #NONE

#Relationship between Water and Amoeba
all.equal(is.na(otter$Water), is.na(otter$Amoeba)) # 80 element mismatches?

#Some Cleaning

numericColumns = c(2, 3, 4, 6, 9, 10, 11, 12, 13, 14, 15)
categoricalColumns = c(1, 5)
df_colnames = colnames(otter)

transpose_df <- function(df) {
  t_df <- data.table::transpose(df)
  colnames(t_df) <- rownames(df)
  rownames(t_df) <- colnames(df)
  t_df <- t_df %>%
    tibble::rownames_to_column(.data = .) %>%
    tibble::as_tibble(.)
  return(t_df)
}

#Data Dictionary

```

# The Data

The data for the analysis were obtained by International Hugs for Otters (IHO).  The data was collected over the period from July 2018 to February 2020 on five  river  otters  (RO-599,  RO-106, RO-255, RO-918, RO-859) located in five distinct regions on the Island of Isle Royale inLake Superior, north of the state of Michigan’s upper peninsula.  Each observation is aggregated to represent otter/region  characteristics  each  month.  The data contains 100 observations of 15 variables.  Description of all variables can be found in the data dictionary of the appendix.

## Examining Missing Data

From the given data there are 100 observations. However there are 84 (84%) observations that contain one or more missing values.  Of the observations that contain missing values, the missing values were observed for one or multiple of the following variables:  Altitude, Trout, Amoeba, Water.  However, two of the variables with missing observations, Amoeba and Water, are not applicable to some sites, and are therefore marked as "N/A", causing an artifically high censorship rate.  Since there is a systemic cause for missingness in these two variables, we will perform different data processing to adjust for missing values compared to the method used for Altitude and Trout. 


```{r}
# Missingness Patter Plot
#Did I do this right?
#plot = mice::md.pattern(otter %>% select(3, 4, 11, 13),plot = TRUE,rotate.names = TRUE)
#View(plot)
```


#### Data Interpolation ofor Altitute and Trout

The variables representing altitude and trout had 18 and 16 "N/A" observations, respectively. These variables represent the average change in altitude and amount of trout present over the month of observation.  Due to the time series nature of the data, we used an interpolation method to correct the missing data in the set.  Observations with "N/A" for either of these variables were replaced with the average of the measurement for that value for the month prior and after that specific observation.  

```{r}
otter <- otter %>%
  group_by(Site) %>%
  mutate(Alt.Interp = na.approx(Altitude, na.rm=FALSE))   

otter <- otter %>%
  group_by(Site) %>%
  mutate(Trout.Interp = na.approx(Trout, na.rm=FALSE)) 
```

#### Transformation and Imputation of Ameoba and Water Variable

The variables representing amoeba and water both had 40 "N/A" observations, meaning that these variables were not applicable to two of the five sites each.  The amoeba variable, which represents green mussel amoeba in the green mussels, was not applicable to the Gamma and Delta site, due to lack of green mussels in these regions.  The water variable was not applicable to the Omega or Alpha site due to the fact that otters in these sites are coastal otters.  

Both variables for Ameoba and Water represent scaled measures rather than pure measurements.  The amoeba scale ranged from -1 to 1, while the water scale was strictly negative, with a minimal value of -27.4014.  To adjust for missingness in these variables, we created new dummy variables with binary values to represent if the value for amoeba or water was missing (1) for that specific observation or not (0).  Next, we performed a linear transformation of +1 to the amoeba variable such that the adjusted scale ranged from 0 to 2.  Finally, we performed a single variable imputation for the amoeba and water variables, where we replaced "N/A" values with 0 if those variables were not applicable to the site.

```{r}
#Add dummy variable representing if the value was missing (1) or not (0)
otter$AmoebaNA <- ifelse(is.na(otter$Amoeba), 1, 0)
otter$WaterNA <- ifelse(is.na(otter$Water), 1, 0)

#Transformation on Amoeba
otter$Amoeba <- otter$Amoeba +1

# Imputation on Amoeba and Water
otter = otter %>%
  mutate(
    Amoeba = replace_na(Amoeba,0),
    Water = replace_na(Water,0)
  )
```

## Descriptive Statistics

```{r}
interest.vars <- c("Activity", "Caloric", "Water.Excess", "Amoeba", "Land", "Water")
descriptives.df <- stat.desc(otter[(interest.vars)], basic = F)
descriptives.df <- descriptives.df[-c(3, 4, 7),]


library(kableExtra)
library(gridExtra)

descriptives.df %>%
kbl(caption = "Descriptive Statistics for Quantitative Variables", label = "Table 1") %>% kable_paper("hover", full_width = F)
```

