---
title: "Descriptive Statistics"
author: "Allison McCarty"
date: "4/11/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)
library(readxl)
library(tidyverse)
library(dplyr)
library(pastecs)
library(data.table)
library(formattable)
library(tidyr)
library(mice)
library(tibble)
library(GGally)
library(kableExtra)
library(ggpubr)


# Read in Data
otter <- read.csv("~/Downloads/otter2.csv")

#Check for NA's
sum(apply(otter, 1, function(x){any(is.na(x))})) #84/100 observations have at least one missing value
sum(is.na(otter$Otter)) # NONE
sum(is.na(otter$Activity)) #NONE
sum(is.na(otter$Altitude)) #18 MISSING VALUES
sum(is.na(otter$Trout)) #16 MISSING VALUES
sum(is.na(otter$Site)) #NONE
sum(is.na(otter$Month)) #NONE
sum(is.na(otter$Lat.)) #NONE
sum(is.na(otter$Long.)) # NONE
sum(is.na(otter$Caloric)) #NONE
sum(is.na(otter$Water.Excess)) #NONE
sum(is.na(otter$Amoeba)) #40 MISSING VALUES
sum(is.na(otter$Land)) #NONE
sum(is.na(otter$Water)) #40 MISSING VALUES
sum(is.na(otter$Month.1)) #NONE
sum(is.na(otter$Year)) #NONE

#Relationship between Water and Amoeba
all.equal(is.na(otter$Water), is.na(otter$Amoeba)) # 80 element mismatches?

#Some Cleaning
otter_raw = read.csv("~/Downloads/otter2.csv") %>%
  mutate(
    Altitude = replace_na(Altitude,"Unknown"),
    Trout = replace_na(Trout,"Unkown"),
    Amoeba = replace_na(Amoeba,"Unkown"),
    Water = replace_na(Water,"Unkown"),
    numMissing = rowSums(across(everything(), ~is.na(.)))
  )
numericColumns = c(2, 3, 4, 6, 9, 10, 11, 12, 13, 14, 15)
categoricalColumns = c(1, 5)
df_colnames = colnames(otter_raw)

try_otter <- otter[-c(1,3,4,5, 11, 13)]

transpose_df <- function(df) {
  t_df <- data.table::transpose(df)
  colnames(t_df) <- rownames(df)
  rownames(t_df) <- colnames(df)
  t_df <- t_df %>%
    tibble::rownames_to_column(.data = .) %>%
    tibble::as_tibble(.)
  return(t_df)
}

#Data Dictionary

```

# The Data

The data for the analysis were obtained by International Hugs for Otters (IHO).  The data was collected over the period from July 2018 to February 2020 on five  river  otters  (RO-599,  RO-106, RO-255, RO-918, RO-859) located in five distinct regions on the Island of Isle Royale inLake Superior, north of the state of Michigan’s upper peninsula.  Each observation is aggregated to represent otter/region  characteristics  each  month.  The data contains 100 observations of 15 variables.  Description of all variables can be found in the data dictionary of the appendix.

## Examining Missing Data

From the given data there are 100 observations. However there are 84 (84%) observations that contain one or more missing values.  The rate of missing observations is high because two of the variables do not apply to some of the otter populations, therefore, those observations were noted as "N/A".  Of the observations that contain missing values, the missing values were observed for one or multiple of the following variables:  Altitude, Trout, Amoeba, Water ** reword***


IDEA: Imputation of Altitude to the median value for each of the sites

```{r}
#Did I do this right?
plot = mice::md.pattern(otter_raw %>% select(3, 4, 11, 13),plot = TRUE,rotate.names = TRUE)
View(plot)
```

## Imputing Continous Variable Representing Trout

After determining that the four variables containing missing values (Altitude, Trout, Amoeba, Water) were all missing at random, we decided to address this issue by performing a single imputation of the median to replace the missing values.

NOTES: **ALLI CLEAN THIS UP**

The first variable with missing data is the variable representing a scaled measurement of lake trout population, which is a major food source for otters.  There were 16 observations with missing trout measurements.  For each of these, we performed a single imputation to replace the missing observation with the median trout value for the specific site. 

```{r}
sites <- c("Alpha", "Delta", "Gamma", "Omega", "Omicron")
filter.otter  <- function(x) {filter(otter, Site == x)}
site.df <- lapply(x=sites, filter.otter)
medianTrout = median(otter$Trout, na.rm = TRUE)
imputedData = raw_data %>%
  mutate(
    balance = replace_na(balance,medianBalance),
    campaign = replace_na(campaign,medianCampaign),
    age = replace_na(age,medianAge),
    duration = replace_na(duration,medianDuration)
  )
```

## Interpolation of Altitude

# Transformation of Ameoba and Water Variable

## Descriptive Statistics

```{r}
mean_df = otter %>%
  summarize(across(numericColumns,function(x) round(mean(x, na.rm = TRUE),2))) %>%
  pivot_longer(everything(), names_to = "Variable",values_to = "mean") %>%
  na.omit()
sd_df = otter %>%
  summarize(across(numericColumns,function(x) round(sd(x, na.rm = TRUE),2))) %>%
  pivot_longer(everything(),names_to = "Variable",values_to = "sd") %>%
  na.omit()
median_df = otter %>%
  summarize(across(numericColumns,function(x) round(median(x,na.rm = TRUE), 2))) %>%
  pivot_longer(everything(),names_to = "Variable", values_to = "median") %>%
  na.omit()
pecent_missing = otter %>%
  summarize(across(numericColumns,function(x) round((sum(is.na(x))/nrow(otter))*100,2))) %>%
  pivot_longer(everything(),names_to = "Variable", values_to = "Missing Number") %>%
  na.omit()
display = inner_join(mean_df,sd_df) %>%
  inner_join(median_df) %>%
  inner_join(pecent_missing)

ddf = transpose_df(display)
ddf
names(ddf) <- c("Variable","Activity","Altitude","Trout","Month","Caloric","Water.Excess","Amoeb","Land","Water","Month.1","Year")
ddf = ddf[-1,]

library(kableExtra)
library(gridExtra)
ddf %>%
kbl(caption = "Descriptive Statistics for Quantitative Variables", label = "Table 1") %>% kable_paper("hover", full_width = F)
```
