---
title: "Analysis of Environmental and Man-Made Factors Affecting Otter Populations on Isle Royale"
author: "Allison McCarty, Jordan Majoros, Shelley Rao"
date: "4/20/2021"
output: html_document
---

# Abstract (save for the end)

- Describe otter population on Island of Isle Royale
- Frame environmental concerns including the dog spa, and describe potential affect on otter
- Our goal is to perform an anlaysis on trends of the environmental factors (briefly touch on specific aims)
- Summarize results in 1-2 sentences


# Intoduction (edits at the end to include results)

Otters are a popular species that can be found on Michigan's Upper Peninsula in the island of Isle Royale in Lake Superior. In North America, otters are classified as a keystone species.  This distinction indicates that otters have a disproportionately large impact on their community. Because keystone species play a critical role in maintaining the structure and biodiversity within an ecosystem, it is crucial to understand the species and ensure their conservation. 

Otters, like many other species on Isle Royale and worldwide, are vulnerable to the climate change crisis that has been unfolding for many decades.  Fortunately, the otter is not currently on The International Union for Conservation of Nature (IUCN) Red List for endangered species. However, the conservation of these animals is still an urgent issue, as hunting and other human disturbances have nearly driven otters to extinction many times throughout the last two centuries.  In addition, a luxury dog spa on Isle Royale recently opened on Isle Royale. The spa could potentially threaten the well-being of otters on the island, and poses significant environmental concerns relating to unsustainability, deforestation, waste management, water contamination, and energy consumption. The habitat intrusion of caused by the dog spa combine with already dire concerns related to the global climate crisis have created the urgent need to understand ecological factors that impact otters and their ability to survive and thrive on Isle Royale. Thus, International Hugs for Otters (IHO) has partnered with our firm to conduct an analysis on the cyclical, seasonal, and environmental factors that impact the otters' behaviors and activity, and the potential impact of the dog spa on these factors. IHO conducted a 20-month study on five otter rafts on Isle Royale. During the course of the investigation, **we will do something**.  This will include significant data processing and exploratory data analysis. We outlined three specific aims for the purpose of this analysis:

    1. Determine the cyclic, seasonal, and/or environmental factors that impact each otter’s activity level, taking into account the shared ecology, resources, and encounters.

    2. Determine the relationship between food sources and otter activity.

    3. Determine the effect of the luxury dog spa (?) add this after SAP review.
    
A brief paragraph about how we did this, as well as more detail about the plan in the above paragraph

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)
library(readxl)
library(tidyverse)
library(dplyr)
library(pastecs)
library(data.table)
library(formattable)
library(tidyr)
library(mice)
library(tibble)
library(GGally)
library(kableExtra)
library(ggpubr)
library(zoo)
library(gridExtra)


# Read in Data
otter <- read_csv("otter2.csv")


#Some Cleaning
numericColumns = c(2, 3, 4, 6, 9, 10, 11, 12, 13, 14, 15)
categoricalColumns = c(1, 5)
df_colnames = colnames(otter)

transpose_df <- function(df) {
  t_df <- data.table::transpose(df)
  colnames(t_df) <- rownames(df)
  rownames(t_df) <- colnames(df)
  t_df <- t_df %>%
    tibble::rownames_to_column(.data = .) %>%
    tibble::as_tibble(.)
  return(t_df)
}
```


# Data

The original dataset collected by International Hugs for Otters (IHO) contained 100 observations of 15 variables, including two factored categorical variables, two variables representing latitude/longitude location coordinates, and eleven numerical variables.  The data was collected over the period from July 2018 to February 2020 on five river otters  (RO-599,  RO-106, RO-255, RO-918, RO-859) located in five distinct regions on the Island of Isle Royale inLake Superior, north of the state of Michigan’s upper peninsula. Each observation is aggregated to represent otter/region  characteristics  each  month.  There are a total of 20 observations for each of the 5 sites, representing a particilar variable for a given month in the 20-month study.  Description of all variables can be found in the data dictionary of the appendix.

## Examining Missing Data

From the given data there are 100 observations. However there are 84 (84%) observations that contain one or more missing values.  Of the observations that contain missing values, the missing values were observed for one or multiple of the following variables:  Altitude, Trout, Amoeba, Water. However, two of the variables with missing observations, Amoeba and Water, are not applicable to some sites, and are therefore missing, causing an artifically high censorship rate.  Since there is a systemic cause for missingness in these two variables, we will perform different data processing to adjust for missing values compared to the method used for Altitude and Trout. 

- **ALLI ADD HERE: Paragraph explaining how we are investigating patterns of missingness**


```{r}
library(naniar)
otter.missing <- otter[,-c(7, 8, 15)]
miss1<- gg_miss_fct(x = otter.missing, fct = Month) + labs(title = "Missingness Pattern by Month")
miss2 <- gg_miss_fct(x = otter.missing, fct = Otter) + labs(title = "Missingness Pattern by Otter")
ggarrange(miss1, miss2, ncol=1)
```


#### Data Interpolation for Altitute and Trout

The variables representing altitude and trout had 18 and 16 missing values, respectively. These variables represent the average change in altitude and amount of trout present over the month of observation.  Due to the time series nature of the data, we used an interpolation method to correct the missing data in the set.  Missing values for either of these variables were replaced with the average of the measurement for that value for the month prior and after that specific observation. 

#### Transformation and Imputation of Ameoba and Water Variable

The variables representing amoeba and water both had 40 missing values, meaning that these variables were not applicable to two of the five sites each.  The amoeba variable, which represents green mussel amoeba in the green mussels, was not applicable to the Gamma and Delta site, due to lack of green mussels in these regions.  The water variable was not applicable to the Omega or Alpha site due to the fact that otters in these sites are coastal otters.  

Both variables for Ameoba and Water represent scaled measures rather than pure measurements.  The amoeba scale ranged from -1 to 1, while the water scale was strictly negative, with a minimal value of -27.4014.  To adjust for missingness in these variables, we created new dummy variables with binary values to represent if the value for amoeba or water was missing (1) for that specific observation or not (0).  Next, we performed a linear transformation of +1 to the amoeba variable such that the adjusted scale ranged from 0 to 2.  Finally, we performed a single variable imputation for the amoeba and water variables, where we replaced missing values with 0 if those variables were not applicable to the site. The final data set contains 100 total observations of 17 variables with 0 missing values. Descriptive statistics for the final dataset is included in Table X in the appendix.

```{r}
otter <- otter %>%
  group_by(Site) %>%
  mutate(Alt.Interp = na.approx(Altitude, na.rm=FALSE))   

otter <- otter %>%
  group_by(Site) %>%
  mutate(Trout.Interp = na.approx(Trout, na.rm=FALSE)) 
```


```{r}
#Add dummy variable representing if the value was missing (1) or not (0)
otter$AmoebaNA <- ifelse(is.na(otter$Amoeba), 1, 0)
otter$WaterNA <- ifelse(is.na(otter$Water), 1, 0)

#Transformation on Amoeba
otter$Amoeba <- otter$Amoeba +1

# Imputation on Amoeba and Water
otter = otter %>%
  mutate(
    Amoeba = replace_na(Amoeba,0),
    Water = replace_na(Water,0)
  )
```


# Exploratory Data Analysis

First, we will check the normality assumption and the distribution plots for all of the variables. We made QQPlots for all of the continuous numeric variables, which is included as Figure X in the appendix.  Based on the plots, we think that caloric intake, water excess, and altitude are approximately normal (at least normal enough that they do not merit transformations).

Based on the QQPlots, variables representing activity, trout, amoeba, land, and water deviate from the normal distribution assumption. We will take a closer look at these variables and try a log transformation if necessary.


```{r}
otter.abnormal <- otter.cont[c(1, 3, 6:8)]
library(reshape2)
otter.abnormal<- melt(otter.abnormal)
ggplot(data = otter.abnormal , aes(x = value)) + stat_density() + facet_wrap(~variable, scales = "free")
#Trout, Activity, and Land have a strong right skew, whereas water has a left skew.  However, all water variables are negative, so I am not sure how to deal with them

#Now, let's take a log transformation of all of these variables
ggplot(data = otter.abnormal , aes(x = log(value))) + stat_density() + facet_wrap(~variable, scales = "free")
```

Pairs Plot : could be cleaned up a bit
```{r}
interestcols <- c(2, 9, 10:11, 16:17)
ggpairs(otter, columns = interestcols, ggplot2::aes(colour= Site, alpha=.05)) 
```


Let's look at each of the otters through time
```{r}
#New data frames to select variables specific to each site
RO599 <- filter(otter[c(1, numericColumns)], Otter=="RO-599")
RO106 <- filter(otter[c(1, numericColumns)], Otter=="RO-106")
RO255 <- filter(otter[c(1, numericColumns)], Otter=="RO-255")
RO918 <- filter(otter[c(1, numericColumns)], Otter=="RO-918")
RO859 <- filter(otter[c(1, numericColumns)], Otter=="RO-859")

# Activity, Altitude, Trout, Caloric, Water

p1 <- ggplot(data= otter, aes(y=Activity, x=Month, color=Site))+geom_smooth(se=FALSE)
p2 <- ggplot(data= otter, aes(y=Altitude, x=Month, color=Site))+geom_smooth(se=FALSE)
p3 <- ggplot(data= otter, aes(y=Trout, x=Month, color=Site))+geom_smooth(se=FALSE)
p4 <- ggplot(data= otter, aes(y=Caloric, x=Month, color=Site))+geom_smooth(se=FALSE)
p5<- ggplot(data= otter, aes(y=Water, x=Month, color=Site))+geom_smooth(se=FALSE)
p6 <- ggplot(data= otter, aes(y=Amoeba, x=Month, color=Site))+geom_smooth(se=FALSE)
ggarrange(p1, p2, p3, p4, p5, p6)
```


## Appendix

```{r}
#Data Dictionary
library(readxl)
library(kableExtra)
library(gridExtra)
data_dict = read_excel("Data_dictionary.xlsx")
data_dict %>%
kbl(caption = "Data Dictionary", label = "Table 1") %>% kable_paper("hover", full_width = F) %>% kable_styling(full_width = TRUE)

#Descriptive Statistics
interest.vars <- c("Activity", "Caloric", "Water Excess", "Amoeba", "Land", "Water")
descriptives.df <- stat.desc(otter[(interest.vars)], basic = F)
descriptives.df <- descriptives.df[-c(3, 4, 7),]
descriptives.df %>% mutate_if(is.numeric, round, digits=2) %>%
kbl(caption = "Descriptive Statistics for Quantitative Variables", label = "Table 2") %>% kable_paper("hover", full_width = F)

#QQNorm Plots
library(car)
otter.cont <- otter[numericColumns]
otter.cont <- otter.cont[,-c(4, 10, 11)]
cont.labels <- colnames(otter.cont)
newfun <- function(x, y) qqPlot(x, main = paste(y, "QQ Plot", sep = " "))
par(mfrow = c(2, 3))
mapply(newfun, otter.cont, cont.labels)
```

