---
title: Analysis of Environmental and Man-Made Factors Affecting Otter Populations
  on Isle Royale
author: "Allison McCarty, Jordan Majoros, Shelley Rao"
date: "4/20/2021"
output: pdf_document
---

# Abstract (save for the end)

- Describe otter population on Island of Isle Royale
- Frame environmental concerns including the dog spa, and describe potential affect on otter
- Our goal is to perform an anlaysis on trends of the environmental factors (briefly touch on specific aims)
- Summarize results in 1-2 sentences


# Introduction (edits at the end to include results)

Otters are a popular species that can be found on Michigan's Upper Peninsula in the island of Isle Royale in Lake Superior. In North America, otters are classified as a keystone species.  This distinction indicates that otters have a disproportionately large impact on their community. Because keystone species play a critical role in maintaining the structure and biodiversity within an ecosystem, it is crucial to understand this species and ensure their conservation. 

Otters, like many other species on Isle Royale and worldwide, are vulnerable to the climate change crisis that has been unfolding for many decades.  Fortunately, the otter is not currently on The International Union for Conservation of Nature (IUCN) Red List for endangered species. However, the conservation of these animals is still an urgent issue, as hunting and other human disturbances have nearly driven otters to extinction many times throughout the last two centuries.  In addition, a luxury dog spa on Isle Royale recently opened on Isle Royale. The spa could potentially threaten the well-being of otters on the island, and poses significant environmental concerns relating to unsustainability, deforestation, waste management, water contamination, and energy consumption. The habitat intrusion of caused by the dog spa combine with already dire concerns related to the global climate crisis have created the urgent need to understand ecological factors that impact otters and their ability to survive and thrive on Isle Royale. Thus, International Hugs for Otters (IHO) has partnered with our firm to conduct an analysis on the cyclical, seasonal, and environmental factors that impact the otters' behaviors and activity, and the potential impact of the dog spa on these factors. IHO conducted a 20-month study on five otter rafts on Isle Royale. During the course of the investigation, perform a statistical analysis considering observed seasonal trends in the context of climate change and the recent opening of the luxury dog spa.  This analysis will include significant data processing and exploratory data analysis. We outlined two specific aims for the purpose of this analysis:

   1. Determine the cyclic, seasonal, and/or environmental factors that impact each otter’s activity level, taking into account the shared ecology, resources, and encounters.

  2. Determine the relationship between food sources and otter activity.
    
- A brief paragraph about how we did this, as well as more detail about the plan in the above paragraph

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)
library(readxl)
library(tidyverse)
library(dplyr)
library(pastecs)
library(data.table)
library(formattable)
library(tidyr)
library(mice)
library(tibble)
library(GGally)
library(kableExtra)
library(ggpubr)
library(zoo)
library(gridExtra)
library(readxl)
library(car)

# Read in Data
otter <- read_csv("otter2.csv")
otter.before <- otter

#Some Cleaning
numericColumns = c(2, 3, 4, 6, 9, 10, 11, 12, 13, 14, 15)
categoricalColumns = c(1, 5)
df_colnames = colnames(otter)

transpose_df <- function(df) {
  t_df <- data.table::transpose(df)
  colnames(t_df) <- rownames(df)
  rownames(t_df) <- colnames(df)
  t_df <- t_df %>%
    tibble::rownames_to_column(.data = .) %>%
    tibble::as_tibble(.)
  return(t_df)
}
```


# Data

The original dataset collected by International Hugs for Otters (IHO) contained 100 observations of 15 variables, including two factored categorical variables, two variables representing latitude/longitude location coordinates, and eleven numerical variables.  The data were collected over the period from July 2018 to February 2020 on five river otters (RO-599, RO-106, RO-255, RO-918, RO-859) located at five distinct sites on the Island of Isle Royale in Lake Superior, north of the state of Michigan’s upper peninsula. Each observation is aggregated to represent otter characteristics each month.  There are a total of 20 observations for each of the 5 sites, where each observation represents a particular variable for a given month in the 20-month study.  A description of all variables can be found in the data dictionary of the appendix.

### Examining Missing Data

From the given data there are 100 observations. However, there are 84 (84%) observations that contain one or more missing value.  Of the observations that contain missing values, the missing values were observed for one or multiple of the following variables:  altitude, trout, amoeba, water. However, two of the variables with missing observations, amoeba and water, are not applicable to some sites, and are therefore missing for all observations at some sites, causing an artificially high censorship rate.  

The amoeba variable represents a scaled measure of the presence of green mussel amoeba in green mussels, which is an important food source for some otters.  However, green mussels are not present in two of the five regions of the study and are therefore not applicable to two of the observed otter populations (RO-255 and RO-106).  The water variable represents a scaled measure of inland water availability.  This metric is not applicable to costal otters and therefore not observed in two of the otter populations in this study (RO-599 and RO-859).  

To investigate the missingness pattern of the data, we constructed a missingness plot in Figure X.  The plot maps missingness by month and otter raft, where the color of the graph is scaled to represent the frequency of missingness.  As mentioned, the amoeba and water variables are not applicable to all otter types, so we expect to see a distinct missingness pattern. Missingness of these variables is dependent on otter type, as demonstrated by the yellow bars in the bottom panel of Figure X.  On the other hand, missingness in variables representing altitude and trout are relatively infrequent in all otter rafts and at all time points, which suggests that data for these two variables are missing at random. Since there is a systemic cause for missingness in these two variables, we will perform different data processing to adjust for missing values compared to the method used for altitude and trout.


```{r, fig.align='center', fig.cap="Missingness Plot of Variables by Month and Otter"}
library(naniar)
otter.missing <- otter[,-c(7, 8, 15)]
miss1<- gg_miss_fct(x = otter.missing, fct = Month) + labs(title = "Missingness Pattern by Month")
miss2 <- gg_miss_fct(x = otter.missing, fct = Otter) + labs(title = "Missingness Pattern by Otter")
ggarrange(miss1, miss2, ncol=1)
```

#### Transformation and Imputation of Amoeba and Water Variable

Amoeba and water were not applicable to two of the five sites, meaning that these variables had 40 missing values each. Both variables for amoeba and water represent scaled measures rather than pure measurements.  The amoeba scale ranged from -1 to 1, while the water scale was strictly negative, with a minimal value of -27.4014.  To distinguish whether these variables were relevant to a particular site, we created new dummy variables with binary values (Yes/No) to represent if the value for amoeba or water was missing from the original dataset. Next, we performed a linear transformation of +1 to the amoeba variable such that the adjusted scale was strictly positive on a range from 0 to 2.  Finally, we performed a single variable imputation for the amoeba and water variables, where we replaced missing values with 0 if those variables were not applicable to the site. 


#### Data Interpolation and Extrapolation for Altitude and Trout 

The variables representing altitude and trout had 18 and 16 missing values, respectively. We determined above that these values were missing at random.  These variables represent the average change in altitude and amount of trout present over the month of observation.  As the data is aggregated to represent scaled measurements over certain months, we wanted to address missing observations such that seasonal patterns and trends are observed. Therefore, we used an interpolation method to correct the missing data in the set for these two variables.  Using the interpolation method, missing values for either of these variables were replaced with the average of the measurement for that value for the month prior and after that specific observation. However, the interpolation method does not work for observations in the first of last month of the study, as average values cannot be found without having information on both the moth prior and the month after a specific observation.  For missing values in either the first or last month, an extrapolation method was used instead.  Using the extrapolation method, missing values in the first month of the study were replaced with the value of that observation in the second month, whereas missing values in the last month (Month 20) were replaced with the value of that observation in the second-to-last month (Month 19).

Figure X shows overlapping density plots for trout and altitude variables before and after the interpolation and extrapolation methods were applied to replace missing values.  The plots show significant overlaps with minor shifts in the density distribution after missing data was filled in.  Descriptive statistics for altitude and trout before and after data processing are also included in Table X.  There are some discrepancies in mean, standard deviation, and variances of the distributions.  To confirm that these discrepancies were not significant, we performed a two-sample t-test for differences in the mean for both altitude and trout before and after data processing.  We obtained p-values of 0.9663 and 0.7021 for the comparison of before and after data processing in these variables, respectively, confirming that the data processing did not cause a significant change in the distribution about the mean.

The final data set contains 100 total observations of 17 variables with 0 missing values. Descriptive statistics for the final dataset are included in Table X in the appendix.


```{r}
otter <- otter %>%
  group_by(Site) %>%
  mutate(Alt.Interp = na.approx(Altitude, na.rm=FALSE, rule=2))
# Make sure that they all were covered
# sum(is.na(otter$Alt.Interp))
otter <- otter %>%
  group_by(Site) %>%
  mutate(Trout.Interp = na.approx(Trout, na.rm=FALSE, rule=2))
# sum(is.na(otter$Trout.Interp))
```


```{r}
#Add dummy variable representing if the value was missing (1) or not (0)
otter$AmoebaNA <- ifelse(is.na(otter$Amoeba), 1, 0)
otter$WaterNA <- ifelse(is.na(otter$Water), 1, 0)

#Transformation on Amoeba
otter$Amoeba <- otter$Amoeba +1

# Imputation on Amoeba and Water
otter = otter %>%
  mutate(
    Amoeba = replace_na(Amoeba,0),
    Water = replace_na(Water,0)
  )
```

```{r}
interest.vars <- c("Trout", "Altitude")
interest.vars2 <- c("Trout.Interp", "Alt.Interp")
after <- stat.desc(otter[(interest.vars2)], basic = F)
before <- stat.desc(otter.before[(interest.vars)], basic = F)

descriptives.total <- cbind(before, after)
descriptives.total <- descriptives.total[-c(1, 3, 4, 7),]

descriptives.total %>% mutate_if(is.numeric, round, digits=2) %>%
kbl(caption = "Descriptive Statistics for Quantitative Variables Before Data Processing", label = "Table 2") %>% kable_classic_2(full_width = F)
```


```{r, fig.align='center', fig.cap="Density Distributions for Altitude and Trout Before and After Data Transformation"}
#Sample data
Altitude <- c(otter$Altitude, otter$Alt.Interp)
Trout <- c(otter$Trout, otter$Trout.Interp)
Transform <- c(rep("Original", 100), rep("Transformed", 100))
OverlapData <- data.frame(Altitude, Trout, Transform)
#Plot
o1 <- ggplot(OverlapData, aes(x = Altitude, fill = Transform)) + geom_density(alpha = 0.5)
o2 <- ggplot(OverlapData, aes(x = Trout, fill = Transform)) + geom_density(alpha = 0.5)
ggarrange(o1, o2, common.legend = TRUE)
```


```{r, include=FALSE}
# t-test for difference in sample means
t.test(otter$Altitude, otter$Alt.Interp, paired = FALSE, conf.level = .95, alternative = "two.sided")
t.test(otter$Trout, otter$Trout.Interp, paired = FALSE, conf.level = .95, alternative = "two.sided")
```



# Exploratory Data Analysis

First, we will check the normality assumption and the distribution plots for all of the variables. We made QQPlots for all of the continuous numeric variables, which is included as Figure X in the appendix.  Based on the plots, we think that caloric intake, water excess, and altitude are approximately normal (at least normal enough that they do not merit transformations).

Based on the QQPlots, variables representing activity, trout, amoeba, land, and water deviate from the normal distribution assumption. We will take a closer look at these variables and try a log transformation if necessary.

```{r, include=FALSE}
#QQNorm Plots
library(car)
otter.cont <- otter[numericColumns]
otter.cont <- otter.cont[,-c(4, 10, 11)]
cont.labels <- colnames(otter.cont)
newfun <- function(x, y) qqPlot(x, main = paste(y, "QQ Plot", sep = " "))
par(mfrow = c(2, 3))
mapply(newfun, otter.cont, cont.labels)
```


We do not expect that Water and Amoeba will be normal based on the way that we filled in the missing data.  Based on the plot, most are normal.  

```{r , fig.align='center', fig.cap="Density Distributions for Continuous Variables"}
otter.abnormal <- otter.cont[c(1, 3, 7)]
library(reshape2)
otter.abnormal<- melt(otter.abnormal)
ggplot(data = otter.abnormal , aes(x = value)) + stat_density() + facet_wrap(~variable, scales = "free")
# We could take a closer look at Activity, Trout, and Land.

#Now, let's take a transformation of all of these variables
ggplot(data = otter.abnormal , aes(x = log(value))) + stat_density() + facet_wrap(~variable, scales = "free")
ggplot(data = otter.abnormal , aes(x = sqrt(value))) + stat_density() + facet_wrap(~variable, scales = "free")
ggplot(data = otter.abnormal , aes(x = (value)^(1/3))) + stat_density() + facet_wrap(~variable, scales = "free")

# do not include all of these graphs, only the transofrmations that we actually perform, all on one plot
# Use the log transformation of Trout, and Land not transform Activity due to interpretability of the outcome variable -- ALLI to explain this in text ("sometimes you're just given bad data"  -Spencer)
ggplot(data=otter, aes(x=Activity))+ stat_density()
ggplot(data=otter, aes(x=log(Activity)))+ stat_density()
ggplot(data=otter, aes(x=(Activity)^1/2))+ stat_density()
```

- Add explanation here about which transformation we choose for the vars (could choose any of the three for any var, does not have to be the same transformation, obviously), and actually perform the transformation to the data in the dataset

Pairs Plot
```{r, fig.align='center', fig.cap="Association Matrix for the Collinearity of Continuous Variables"}
interestcols <- c(2, 9, 12, 17)
ggpairs(otter, columns = interestcols, ggplot2::aes(colour= Site, alpha=.05)) 
ggplot(data=otter, aes(x=Caloric, y=log(Land)))+geom_point()
```

In the pairs plot, we notice the relationships among each of the continuous variables. Caloric and Land have the strongest relationships, with a negative linear trend and strong correlation. Otherwise, the data appears to not have specific trends among the otters for the other variables, with weaker correlations and no evidence of linear trends. 

Alternative Pairs Plot
```{r}
pairs(~Activity + Caloric +  `Water Excess`  + Amoeba + Alt.Interp + Trout.Interp, data=otter, upper.panel=NULL)
```

Let's look at each of the otters through time
```{r, fig.align='center', fig.cap="Patterns of Variables Through Time"}
#New data frames to select variables specific to each site
RO599 <- filter(otter[c(1, numericColumns)], Otter=="RO-599")
RO106 <- filter(otter[c(1, numericColumns)], Otter=="RO-106")
RO255 <- filter(otter[c(1, numericColumns)], Otter=="RO-255")
RO918 <- filter(otter[c(1, numericColumns)], Otter=="RO-918")
RO859 <- filter(otter[c(1, numericColumns)], Otter=="RO-859")

# Activity, Altitude, Trout, Caloric, Water

p1 <- ggplot(data= otter, aes(y=Activity, x=Month, color=Site))+geom_smooth(se=FALSE)
p2 <- ggplot(data= otter, aes(y=Alt.Interp, x=Month, color=Site))+geom_smooth(se=FALSE)
p3 <- ggplot(data= otter, aes(y=Trout.Interp, x=Month, color=Site))+geom_smooth(se=FALSE)
p4 <- ggplot(data= otter, aes(y=Caloric, x=Month, color=Site))+geom_smooth(se=FALSE)
p5<- ggplot(data= otter, aes(y=Water, x=Month, color=Site))+geom_smooth(se=FALSE)
p6 <- ggplot(data= otter, aes(y=Amoeba, x=Month, color=Site))+geom_smooth(se=FALSE)
ggarrange(p1, p2, p3, p4, p5, p6, common.legend = TRUE)


```

As we observe each of the otter's patterns for each variable over time, it is important to note an interesting change at month 10. As seen in the trends with Activity, Trout, and Amoeba, there is a difference in trend which could signify the opening of the dog spa on the island. When the dog spa opened, it would influence the environment around it, potentially harming the otter's food source. Otter Alpha? had a decrease in Amoeba, which signifies a decrease in their food source of mussels around that 10 month mark. On the other hand, Otter Alpha? also increased Activity around the same time, which could signify that it is working harder to find food and exerting more energy. 

1.1.2 Frequency Distribution of each otter and sum of total covered miles

```{r}
total.RO106 <- sum(RO106$Activity)
total.RO255 <- sum(RO255$Activity)
total.RO599 <- sum(RO599$Activity)
total.RO859 <- sum(RO859$Activity)
total.RO918 <- sum(RO918$Activity)

OtterNames <- c("RO106", "RO255", "RO599", "RO859", "RO918")
TotalDistance <- c(total.RO106, total.RO255, total.RO599, total.RO859, total.RO918)
Freq.Dist <- cbind(id=OtterNames, TotalDistance)

Freq.Dist %>%  kbl(caption = "Frequency Distribution for Total Distance Covered by Otter", label = "Table 2") %>% kable_classic_2(full_width = F)
```

1.2.2 and 1.2.3 --> Ask about plan for mapping

1.2.4 Scatterplot Showing the Relationship between Trout and Total Distance Covered
```{r}
ggplot(otter, aes(x = Trout.Interp, y=Activity, color=Otter))+geom_point()+geom_smooth(se=FALSE)
```

Next, we created a scatterplot to examine the relationship between lake trout population and total distance covered. We observe that there is a generally positive association between lake trout population and total distance covered. On average, islands resided by RO-255, RO-859, RO-918 have lower activity levels than islands resided by RO-106 and RO-599. Interestingly, the island resided by RO-599 have a much higher lake trout population (up to 1777), as compared to other islands (up to 800). Moreover, there is a noticeably steeper slope as lake trout population increases which indicates much more activity gains at higher levels of trout. 

1.2.5 Scatterplot Showing the Relationship between amount of Amoeba and Caloric Expenditure

Next, we created a scatterplot to examine the relationship between the presence of the green mussel amoeba in the green mussels and caloric expenditure. We observe that amoeba levels differ across islands, with the island resided by RO-599 having the highest mean levels of amoeba, followed by island resided by RO-859, then island resided by RO-918. Islands resided by otters RO-106 and RO-255 have no presence of green mussel amoeba. The plot shows that there is no apparent relationship between the presence of the green mussel amoeba in the green mussels and caloric expenditure.

```{r}
ggplot(otter, aes(x = Amoeba, y=Caloric, color=Otter))+geom_point()

# summary(RO918$Amoeba) ; Maybe something important to note here is the the minimum for RO-918 is .01, NOT 0, so there is in fact a difference in the way that this is coded versus the sites where this variable is not applicable
```

```{r}
df <- aggregate(otter[, 11], list(otter$Otter), mean)

ggplot(data=df, aes(x=Group.1,y=Amoeba,fill=Group.1)) +  geom_bar(stat="identity") + ggtitle("Average Amoeba Level Across Otter Sites") + xlab("Otter Site") + ylab("Level of Amoeba")
```



1.2.6 Bar Graph Comparing Distance for locations with and without Mussels

```{r}
Mussel<- c("No", "No", "Yes", "Yes", "Yes")
Freq.Dist2 <- data.frame(OtterNames, TotalDistance, Mussel)
ggplot(data=Freq.Dist2, aes(x=OtterNames, y=TotalDistance, fill=Mussel)) +
  geom_bar(stat="identity")
```

Next, we create a bar graph to examine the total distance covered each otter. We observe that RO-599 and RO-106 have the greatest total distance covered by otter (about 12,400 and 8,000 respectively). Interestingly, the presence of mussels does not appear to have any association with total distance covered.

1.2.7 Plot Monthly distance covered by each otter

- Already completed in the important vars v. Month section

Linear mixed models are an extension of simple linear models to allow both fixed and random effects, where non-independence exists in the data. A fixed effect is an unknown constant that we try to estimate from the data. A random effect is a random variable-- here, we are less interested in estimating the random effect and more so, the parameters that describe the distribution of the random effect. Using a random effects approach allows us to try to use our analysis to describe the larger population of otters beyond our sample of five otters. 

To study cyclical, seasonal, and environmental factors associated with otter activity, we used a linear mixed-effects model that was created using all combined otter data. The model accounted for variation across individual otter with a random effect for otter identity; the data is grouped by otter and the random effect is constant within each group. The linear model shows that Month and Trout are significant, where every additional month is associated with 9.18 decrease in activity and every additional unit trout is associated .67 increase in activity. 

The mixed model shows that every additional month is associated with .82 units decrease in activity and every additional unit trout is associated with 0.43 increase in activity. The mean number activity level for all otters is 147.44. From the maximum likelihood estimated computed, we see that the between-subjects SD is 208.67. The 95% confidence intervals plot for random effects shows that no interval lies on 0, which means that random effects meaningful in both directions. For Otter 599 and 106, the random effects is positive and for Otter 918, 859, and 255, the effect is negative. 

```{r}
# fixed effects one-way ANOVA
lmod <- aov(Activity ~ Month + Trout.Interp + AmoebaNA + Caloric, otter)
# Month and Trout is significant
summary(lmod) 
# Overall estimated mean activity is 147.44
coef(lmod)
# 1 indicates random effect is constant within each group of otters
library(lme4)
mmod <- lmer(Activity ~ Month + Trout.Interp + AmoebaNA + Caloric + (1|Otter), data=otter)
summary(mmod, digits=2)
# compute maximum likelihood estimates to determine the parameters of the distribution that best describe the given data (as opposed to pop); biased towards zero
smod <-  lmer(Activity ~ Month + Trout.Interp + AmoebaNA + Caloric + (1|Otter), data=otter, REML=FALSE)
summary(smod)
# predicted random effects
ranef(mmod)$Otter
# predicted fixed effects
(cc <- model.tables(lmod))
# calculate 95% confidence intervals for the random effects
library(lattice)
dotplot(ranef(mmod, condVar=TRUE))
# diagnostics
qqnorm(residuals(mmod),main="")
plot(fitted(mmod),residuals(mmod),xlab="Fitted",ylab="Residuals")
abline(h=0)
```


2.1 For the three site locations with presence of mussels, examine the association between amount of amoeba and combined otter activity using an OLS regression model. The explanatory variables are Month of Collection, Trout Population, Caloric Expenditure, Time Spent on Land, and Amount of Amoeba. The outcome variable is Activity level.

In order to fully understand the impact of the environment on the otters, we believed it best to analyze the food source that is amoeba. An important factor within the otter's well being is their food intake, and it important to understand the influence that this food source has on the otter activity level. We used month of collection, trout population, caloric expenditure, and time spent on land as explanatory variables examine the association with amoeba. 

To best understand the relationship between the amoeba and activity, we created a model that included month, trout, and amoeba as the predictors with activity as the response. We chose month and trout through a model selection process that went through each combination of the variables month, trout, caloric, and land to choose the best combination. 

```{r}
#remove otters 106 and 255, 21:60 - just the ones with amoeba
model.dat = otter[-c(21:60),]
#head(model.dat)
trainIndexes = sample(1:nrow(model.dat), size = nrow(model.dat)/2)
training = model.dat[trainIndexes,]
testing = model.dat[-trainIndexes,]

set.seed(123)
formulas <- gen_formula <- function(picks,curfeatures = c("Month","Trout.Interp","Caloric","Land", "Amoeba")) {
  str_c('Activity ~',str_c(curfeatures[picks], collapse = '+'))
}
cv_model <- function(formula) {
  sample <- training[sample(nrow(training)),]
  folds <- cut(seq(1,nrow(sample)), breaks = 10, labels = FALSE)
  results = array(data = -1, dim = 10)
  for (i in 1:10) {
    testIndexes <- which(folds == i, arr.ind=TRUE)
    testData <- sample[testIndexes,]
    trainData <- sample[-testIndexes,]
    model <- lm(formula, data = trainData)
    prediction = predict(model,testData, type = 'response')
    results[i] = mean(testData$Activity == prediction)
  }
  mean(results)
}
#at least two 
possPicks = list(c(1, 2),c(1, 3),c(1, 4),c(1,5),c(2, 3),c(2, 4),c(2,5),c(3, 4),c(3,5),c(4,5),c(1, 2, 3),c(1,2,5),c(1,3,5),c(1,4,5),c(1, 2, 4),c(2,3,5),c(2,4,5),c(3,4,5),c(1, 3, 4),c(2, 3, 4),c(1, 2, 3, 4),c(1,2,3,4,5))
all_formulas = sapply(possPicks,formulas)
cv_scores = sapply(all_formulas, cv_model)
bestFormula = all_formulas[which.max(cv_scores)]
bestFormula
otter.lm = lm(Activity ~ Month + Trout.Interp + Amoeba, data = model.dat)
```


2.1.1 Perform diagnostic analysis to check the assumptions of constant variance and normality using residual plots, QQ-plots or relevant tests. Potential outliers or influential points will be examined.

```{r}
newtable <- summary(otter.lm)$coefficients
kable(newtable, digits=5)

par(mfrow = c(2, 2))
plot(otter.lm)
```

When observing the diagnostics of the model, the model appears normal in the QQ-plot, as it follows the diagonal line with little curvature or deviation. However, the residuals do not appear to be truly homoskedastic, which could cause some concern?. In a residual vs. fitted plot, it is important to see the values be randomly and equally spread out across the 0 line. The values instead appear to be in a cone shape, fanning out towards the higher values. This indicates some hetersoskedasticity, which could mean the data points with larger influence and weight have more of a pull on the model. 

## Discussion and Conclusion

*relation to my model*

Amoeba and trout appear to have a significant relationship to otter activity. As related to the main food sources, they strongly influence how an otter will behave, as more or less active. This is important to note, as any environmental factors that happen to influence the mussels and their amoeba would be a significant predictor to how the otters will behave from that point. If something were to damage the mussels, the otters would be heavily impacted. 

Future Suggestions
- more otters for more collection 
- track population of otters over time as well, see if there is any impact on population from environmental factors, is anything deathly?

## Appendix

```{r}
#Data Dictionary
data_dict = read_excel("Data_dictionary.xlsx")
data_dict %>%
kbl(caption = "Data Dictionary", label = "Table 1") %>% kable_classic_2(full_width = F)

#Descriptive Statistics
interest.vars <- c("Activity", "Caloric", "Water Excess", "Amoeba", "Land", "Water")
descriptives.df <- stat.desc(otter[(interest.vars)], basic = F)
descriptives.df <- descriptives.df[-c(3, 4, 7),]
descriptives.df %>% mutate_if(is.numeric, round, digits=2) %>%
kbl(caption = "Descriptive Statistics for Quantitative Variables", label = "Table 2") %>% kable_classic_2(full_width = F)

#QQNorm Plots
otter.cont <- otter[numericColumns]
otter.cont <- otter.cont[,-c(4, 10, 11)]
cont.labels <- colnames(otter.cont)
newfun <- function(x, y) qqPlot(x, main = paste(y, "QQ Plot", sep = " "))
par(mfrow = c(2, 3))
mapply(newfun, otter.cont, cont.labels)
```

